---
title: "Rmarkdown template"
author: "Thomas Vigié"
output:
  beamer_presentation:
    includes:
      in_header: preamble.txt
    keep_tex: yes
  slidy_presentation: default
classoption: aspectratio=169
urlcolor: blue
linkcolor: SFUblue
---

```{r, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
list.of.packages <- c("tidyverse","rmarkdown")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
# if(length(new.packages)) install.packages(new.packages)
library(tidyverse)
library(rmarkdown)
# library(nycflights13)
# library(lubridate)
# library(Lock5Data)
# library(crimedata)
# library(fivethirtyeight)
# library(ISLR)
# library(stargazer)
# library(randomForest)
# library(party)
# library(tree)
# library(rpart)       # performing regression trees
# library(rpart.plot)  # plotting regression trees
# library(np)
# library(car)
# library(modelr)
# library(rdd)
```

## Disclaimer
I do not allow this content to be published without my consent.


All rights reserved \textcopyright  2023 Thomas Vigié

## Outline  
\label{outline}

- \hyperlink{R}{\textbf{What is \textsf{R}?} }
- \hyperlink{RStudio}{\textbf{What is \textsf{RStudio}?} }
- \hyperlink{script}{\textbf{Editing a script} }
<!-- - \hyperlink{RNG}{\textbf{Generating random numbers (RNG)} } -->
- \hyperlink{R Markdown}{\textbf{What is \textsf{R Markdown}?} }
- \hyperlink{data}{\textbf{Working with data} }
    - \hyperlink{plot}{\textbf{Plots in \textsf{R}} }
    - \hyperlink{commands}{\textbf{Useful commands} }
- \hyperlink{loop}{\textbf{Loops in \textsf{R}} }

## What is \textbf{\textsf{R}}?  
\label{R}

- \textbf{\textsf{R}} is a computational language (Open source!)
- \textbf{\textsf{RStudio}} is a software that provides the interface for \textbf{\textsf{R}} commands
- Can be used to:
    - Analyze data (more tools than Excel)
    - Export data after making operations on them
    - Produce great figures, graphs and tables (nicer than Excel)
    - Perform mathematical operations (fancier than Excel)
    - Write pdf, html, presentation documents, and many more via \textbf{\textsf{R Markdown}}
- You need to download both! \textbf{\textsf{R}} \href{https://www.r-project.org/}{\textbf{here}} first and then \textbf{\textsf{RStudio}} \href{https://www.rstudio.com/}{\textbf{here}} 
    
## How does it work?
- Write code (=instructions) in the \textbf{\textsf{R}} language
- Instructions are sent to the software (run/compile the code)
- \textbf{\textsf{R}} returns an output
- The output can be saved in various formats:
    - **.\textbf{\textsf{R}}Data** (e.g. computations results, data sets)
    - **.csv** (e.g. data sets. What I recommend in practice)
    
## Learning \textbf{\textsf{R}}
- Many books and online resources
- \href{https://swirlstats.com/}{\textbf{Swirl}} (online step by step tutorial. Try the first couple of lessons!)
- \href{https://www.RStudio.com/resources/cheatsheets/}{\textbf{\textsf{R} cheat sheets}}
- \href{https://r4ds.had.co.nz/}{\textbf{\textsf{R} for data science}} is a free online book that teaches data visualization, manipulation, etc... The author Hadley Wickham made huge contributions to the \textbf{\textsf{R}} community, including this book (the book itself was made in \textbf{\textsf{R}}!!)
- Like many languages have the same roots or structures (e.g. French/Spanish/Italian), the \textbf{\textsf{R}} language is very close to other languages like Matlab, Julia, Python. So it is a good investment no matter what
- Knowing at least one computational language is a big (if not the biggest) asset of a young economist these days

## What is \textbf{\textsf{RStudio}}? 
\label{RStudio}

- \textbf{\textsf{R}} is the language, but the interface is not very user friendly...
- \textbf{\textsf{RStudio}} is an **Integrated Development Environment (IDE)** with a waaaay better interface
- It is composed of 4 panels:
    - Top left: This is your **script**, what you edit and save for later
    - Top right: This is the **working environment**. Saved variables, vectors, data sets appear on that panel
    - Bottom left: **The command window**. where the results appear. You can also type commands in it and run them directly (they are not saved in the script in this case)
    - Bottom right: Help files, plots, and working folders can be found among other things

<!-- ## \textbf{\textsf{RStudio}} interface -->
<!-- \begin{center} -->
<!-- \includegraphics[width=\textwidth]{RStudio interface.png} -->
<!-- \end{center} -->


## \textbf{\textsf{RStudio}} interface
\begin{center}
\includegraphics[width=\textwidth]{RStudio interface clear.png}
\end{center}

## Workflow with \textbf{\textsf{RStudio}}
- Have a folder that contains the **.\textbf{\textsf{R}}** or **.\textbf{\textsf{Rmd}}** file, the data, the paper/assignment you are writing and anything related to the script (one project = one folder)
- Open the **.\textbf{\textsf{R}}** or **.\textbf{\textsf{Rmd}}** file by clicking on it every time, it sets the **working directory** directly in that folder (don't skip that step if you want to have a smooth experience. Trust me!!)
- This way, data can be loaded directly, without specifying the whole path of the file
- Believe me, it will avoid you a lot of struggles!

## Workflow with \textbf{\textsf{RStudio}}: Editing a script 
\label{script}

- \textbf{\textsf{R}} contains some functions (= commands), but many commands come in **packages**
- A package is a folder that contains commands for particular purposes
- First thing to do: Install the packages using \texttt{install.packages("package name")} (\textbf{\textsf{RStudio}} will download them from the internet and put them in the appropriate folder automatically)
- Installation can be done once per computer. Every time \textbf{\textsf{RStudio}} is open, you need to **load** the packages using the \texttt{library("package name")} command (\textbf{\textsf{RStudio}} will make the packages ready to use)
- Start editing your script. In general, it involves loading some data set, then performing computations on it and make some comments to keep track of your progress

## Editing a script: Install and load packages

\small
```{r, eval = FALSE, results = "hide", warning = FALSE, message = FALSE}
# This chunk of code takes a list of packages, checks if they are 
# installed already and if not, installs them. 
# Then it loads them and we are ready to go. Credit goes to Chris Muris
list.of.packages <- c("tidyverse","rmarkdown")      
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()
                                   [,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(tidyverse)
library(rmarkdown)

```
\normalsize

## Editing a script: Basic commands
\small
```{r}
print("Hello, world!")    # Print() shows the output inside the brackets
"Ceci n' est pas une pipe" # Or we can make R show it by directly asking 
# for it. Print() is more helpful in general
```

## Editing a script: Basic commands
<!-- \begin{small} -->
```{r}
# You can save the message in the working directory
# (Top right-hand corner) using the <- or = symbol
message <- "Hello World"
message      # that will show what the variable "message" is
 
# Doing some basic statistical operations
data <- rnorm( 20, mean = 5 , sd = 2)  # Creates numbers "randomly", 
# according to a normal distribution
head(data)  # shows the first rows and columns of the data
 
mean(data)
```

## Editing a script: Basic commands
\scriptsize
```{r}
mean(data)  # Computes the sample mean
mean(data^2) # Computes the sample mean of the data with each element squared
 
var(data)                           # variance derivation
sd(data)                            # standard deviation derivation formula
# standard deviation by hand, using the square root. Why are the results different???
sqrt(mean(data^2) - (mean(data))^2) # standard deviation by hand, using the square root
summary(data) # summary() shows a couple of statistics about the data
```
\normalsize
<!-- \end{small} -->


<!-- ## Random number generation (RNG) basics \label{RNG} -->
<!-- - How does a computer generate random number? -->
<!-- - It doesn't. Numbers are generated \textbf{pseudo randomly} -->
<!-- - Idea: start with a number "randomly" picked, perform operations on it, align the result on the distribution it is supposed to come from -->
<!-- - The starting number can come from the computer clock for instance -->

<!-- ## RNG basic algorithm -->
<!-- - One of the oldest algorithm to generate random numbers is the **linear congruential generator** -->
<!-- - Choose a **seed** value (=start value), $X_0$ (can be related to the time displayed by the computer for instance) -->
<!-- - Use the formula -->
<!-- \[ -->
<!-- X_{n+1} = \left(aX_n  + c \right) \text{mod}\, m -->
<!-- \] -->
<!-- where $a \text{mod} b$ means taking the remainder of the division of $a$ by $b$ -->
<!-- - $a$ and $c$ are constants -->
<!-- - For numbers to be generated from a specific distribution, generate a number $U$ between 0 and 1 in a similar way -->
<!-- - Look for the number in the target distribution such that the probability of observing less than it is $U$: It is called the **inversion method** -->


<!-- ## RNG: Illustration -->
<!-- ```{r, eval = FALSE} -->
<!-- n <-   1000              # sample size -->
<!-- dat <- as_tibble(rnorm(n))          # "draw" n numbers from a normal distribution -->
<!-- colnames(dat) <- "x" -->
<!-- g <- ggplot(data = dat) +  # Plot density of x to see the bell curve -->
<!--    geom_density(aes(x = x)) -->
<!-- ``` -->

## What is \textbf{\textsf{R Markdown}}? 
\label{R Markdown}

- \textbf{\textsf{R}} is the language
- \textbf{\textsf{RStudio}} is the interface (the software you use to run everything)    
- \textbf{\textsf{R Markdown}} is a file format for making dynamic documents with \textbf{\textsf{R}}
- We can write plain text, and add code chunks in the middle
- \textbf{\textsf{RStudio}} then compiles the code ("knits") and produces a pdf or html or Word document
- So you can write a paper, a presentation or your resume entirely with \textbf{\textsf{R}} using \textbf{\textsf{R Markdown}}
- The course lecture notes are made in \textbf{\textsf{R Markdown}}
<!-- - It requires loading the ***rmarkdown*** package at the beginning of a session -->

## Workflow with \textbf{\textsf{R Markdown}}
- Write plain text for sections, titles, bullet points etc
- Insert code chunks in between, using ***Ctrl + Alt + I*** (windows)
- Press on the **Knit** button to compile in the format of your choice
- You are **not required** to use \textbf{\textsf{R Markdown}} to produce your paper or presentation, but you are encouraged to try (templates will be made available on cnavas for you to play around)

## \textbf{\textsf{R Markdown}} interface
\begin{center}
\includegraphics[width=\textwidth]{rmarkdown picture.png} -->
\end{center}

## Useful commands and shortcuts in \textbf{\textsf{R Markdown}}
- ***Ctrl + Shift + C*** (windows) or ***Cmd + Shift + C*** (MacOS) turns a line into a comment (same as \textbf{\textsf{RStudio}}, but the comment symbol is not # anymore, it is \<!-- commented text \-->)
- A section starts with #, a subsection with ##, a subsubsection with ### 
- Use the dash **-** to start an itemized (bullet points) list
- ``**'' before and after a word produces bold face
- The dollar sign **$** is used to start the math mode (comes from Latex, an editing programming language). With it, you can type fancy mathematical expressions and symbols such as $\bm{\bar{X}}$, $\bm{X_i}$, $\bm{\sum_{i=1}^{n}\left(y_i - x_i^{\prime}\beta \right)^2}$  
- See the \textbf{\textsf{R Markdown}} cheat sheet \href{https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf}{\textbf{here}} online

## Data analysis in \textbf{\textsf{R}} 
\label{data}

\scriptsize
```{r}
data <- read_csv("vancouver_daily_crime.csv")   # Load the data, and call the data set "data"
names(data)    # See the names of the variables in the data
```
\normalsize

## Data analysis in \textbf{\textsf{R}} 
```{r}
head(data)     # See the first rows of the data
```

## Data analysis in \textbf{\textsf{R}} 
```{r}
summary(data)  # Shows a summary of each variable
```

## Data analysis in \textbf{\textsf{R}} 
```{r}
nrow(data)  # Save the sample size
ncol(data)  # Save the number of variables in the data
# data$Theft    # Looks at one variable in particular
data[6:8, 2]   # looks at the 6th column of the data, 6th to 8th rows
```

## Plots in \textbf{\textsf{R}}: **\texttt{ggplot}**  
\label{plot}

- The **\texttt{tidyverse}** package contains a loooot of features for data analysis
- Highly recommended to always load **\texttt{tidyverse}**
- **\texttt{ggplot}** is one of them. It produces waaay better plots than the basic **\texttt{plot}** command
- It works by layers: First, you tell what data you are using to plot
- Then, you specify the mapping (what $\bm{x}$ and $\bm{y}$ are) using the **\texttt{aes}** command. If you want to assign a color to a third (qualitative) variable, you can use **\texttt{color = }**
- You can change the type of plot using **\texttt{geom\_}** (**\texttt{geom\_point}**, **\texttt{geom\_histogram}**, etc)
- Add more layers to change the way the plot looks: Axis labels, legend style, etc 


## **\texttt{ggplot}** in \textbf{\textsf{R}}: Illustration 
```{r}
mpg <- ggplot2::mpg # This is a data set on cars, 
# contained in the ggplo2 package
names(mpg)    # See the names of the variables in the data
```

## **\texttt{ggplot}** in \textbf{\textsf{R}}: Illustration 
\scriptsize
```{r}
head(mpg)     # See the first row of the data
```

## **\texttt{ggplot}** in \textbf{\textsf{R}}: Illustration 
\scriptsize
```{r}
summary(mpg)  # Shows a summary of each variable
```

## **\texttt{ggplot}** in \textbf{\textsf{R}}: Illustration 
```{r, eval = FALSE}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

## **\texttt{ggplot}** in \textbf{\textsf{R}}: Illustration 

\begin{center}
```{r, echo = FALSE, out.width = '70%'}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy, color = class))
```
\end{center}

## **\texttt{ggplot}** in \textbf{\textsf{R}}: Illustration 
```{r, eval = FALSE}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy, color = class))+
facet_wrap(~ class, nrow = 2)
```

## **\texttt{ggplot}** in \textbf{\textsf{R}}: Illustration 

\begin{center}
```{r, echo = FALSE, out.width = '70%'}
ggplot(data = mpg) +
geom_point(mapping = aes(x = displ, y = hwy, color = class))+
facet_wrap(~ class, nrow = 2)
```
\end{center}


## **\texttt{ggplot}** in \textbf{\textsf{R}}: Illustration 
\scriptsize
```{r}
diamonds <- ggplot2::diamonds  
names(diamonds)    
head(diamonds)     
```
\normalsize

## **\texttt{ggplot}** in \textbf{\textsf{R}}: Illustration 
```{r,  eval = FALSE}
ggplot(data = diamonds) +
geom_bar(mapping = aes(x = cut, fill = clarity))
```

## **\texttt{ggplot}** in \textbf{\textsf{R}}: Illustration 

\begin{center}
```{r,  echo = FALSE, out.width = '70%'}
ggplot(data = diamonds) +
geom_bar(mapping = aes(x = cut, fill = clarity))
```
\end{center}

## Useful commands in \textbf{\textsf{R}} 
\label{commands}

- I will show the implementation of every method covered in the course
- Here are some basic but useful commands:
    - **\texttt{head()}** shows the first lines of a data set
    - **\texttt{summary()}** shows a summary of the data (good to check everything is ok)
    - **\texttt{ggplot()}** for graphs of all kinds (from the **\texttt{ggplot2}** or **\texttt{tidyverse}** packages)
    - **\texttt{nrow()}** and **\texttt{nrow()}** return the number of rows and columns of a data set
    - **\texttt{ifelse()}** creates a 1/0 variable (or other values) based on a condition we specify (1 if the condition is met, 0 otherwise)
    - **\texttt{lm()}** computes the OLS estimator (built-in \textbf{\textsf{R}}. See chapter on linear regression)
    <!-- - **ivreg** computes the 2SLS estimator (***AER*** package) -->
    <!-- - **randomForest** builds a random forest (***randomForest*** package) -->
    <!-- - **rpart** builds a regression or classification tree (***rpart*** package) -->
    <!-- - **svm** builds SVMs (***e1071*** package, and the ***kernlab*** package for svm plots) -->
    <!-- - **hclust** performs clutering (***stats*** package) -->
    <!-- - **prcomp** performs PCA (***stats*** package) -->
- Run **\texttt{??function}** to see help on how to use the command (look for the exact function among the ones proposed on the bottom right panel)
- Example: **\texttt{??lm}**

## Loops in \textbf{\textsf{R}}   
\label{loop}

- A **loop** is a sequence of instructions that the software will execute in an iterated fashion
- 3 types of loops:
    - \textbf{\texttt{for}} loops repeat some instructions a predetermined number of times. Generally, an update is made at the beginning of each iteration
    - \textbf{\texttt{while}} loops execute a sequence of instructions as long as a criterion is satisfied. The criterion is checked at each iteration (so without an update, the loop may never stop!)
    - \textbf{\texttt{if}} "loops" check if a condition is satisfied, execute a command if the condition is **TRUE**, and another command if it is **FALSE**

## Loops in \textbf{\textsf{R}}: Basic \texttt{for} loop
\scriptsize
```{r}
rounds <- 10
vec <- matrix(0, nrow = 1, ncol = rounds) # Making a matrix 
#of 0 that the loop will replace with what I want

for (i in 1:rounds){    # i is the index: It will change at each iteration
vec[1, i] <- i
}
vec
```

## Loops in \textbf{\textsf{R}}: Basic \texttt{while} loop
\scriptsize
```{r}
vec <- c() # Making an empty vector without a specified size
index <- 1
limit <- 15

while (index < limit){    # i is the index: It will change at each iteration
vec[index] <- index*3
index <- index + 1
}
vec
length(vec)   # gives the number of elements in vec
```
## Loops in \textbf{\textsf{R}}: Basic \texttt{if} statements
```{r}
x <- 5
if (x < 8){
  print("x is smaller than 8!!")
  }  else{
  print("x is bigger than 8!!")
  }

# The ifelse function creates an object from  condition to check
y <- ifelse(x >= 5, "x is bigger or equal to 5", "x is strictly smaller than 5")
z <- ifelse(x > 2, x^2, 1)
y
z

```

## Loops in \textbf{\textsf{R}}: A simulation
- Let us make a simulation
- In simulation, we are God: We can generate variables according to the assumptions we want to satisfy
- Simulations are used to see how an estimator does over multiple samples of different size
- We can check the **bias**, **variance**, **consistency** and **distribution** of estimators
- Let us use a loop to generate many samples and compute the mean of each


## Loops in \textbf{\textsf{R}}: A simulation
```{r}
rounds <- 10
vec <- matrix(0, nrow = 1, ncol = rounds) # Making a matrix 
#of 0 that the loop will replace with what I want

for (i in 1:rounds){# i is the index: It will change at each iteration
x <- rnorm(100)
vec[i] <- mean(x)
}
vec
```

## More data to look at
- Many data sets are available through packages. I recommend you check 
    - **\texttt{nycflights13}** 
    - **\texttt{lubridate}** 
    - **\texttt{Lock5Data}** 
    - **\texttt{crimedata}** 
    - **\texttt{fivethirtyeight}**
- Be patient, use online help to learn. You will get a lot of errors, and some of them will be hard to decipher at first 
- It opens the door to many other computer languages
- Have fun!


<!-- ## Something cool -->

<!--  There is a package called `magick` that allows you to include gifs in your plots ! I do not really know how the details work, so I copied and paste the code from this website: \href{http://danielphadley.com/ggplot-Logo/}{\textbf{here}} -->
<!--  See also \href{https://cran.r-project.org/web/packages/magick/vignettes/intro.html}{\textbf{here}} -->

<!-- It takes a gif in the `png` format and adds it to a ggplot. -->

<!-- \scriptsize -->
<!-- ```{r, eval = FALSE} -->
<!--  library(magick) -->
<!--  plot <- image_read("test_error.png") -->
<!--  # And bring in a logo -->
<!--  logo_raw <- image_read("https://i.imgur.com/e1IneGq.jpg") -->
<!--  frames <- lapply(logo_raw, function(frame) { -->
<!--    image_composite(plot, frame, offset = "+50+500") -->
<!--  }) -->

<!--  animation <- image_animate(image_join(frames)) -->
<!--  # animation -->
<!--  # image_write(animation, "test_error_Travolta.gif") -->
<!-- ``` -->

<!-- ## Something cool -->
<!-- \includegraphics{test_error_Travolta.gif} -->
<!-- \movie[label=show3,width=1.0\textwidth,poster -->
<!--        ,autostart,showcontrols,loop] -->
<!--   {\includegraphics[width=1.0\textwidth]{test_error_Travolta.png}}{test_error_Travolta.mp4} -->
<!-- \movie[externalviewer]{\includegraphics[width=\textwidth]{test_error_Travolta.png}}test_error_Travolta.mp4} -->

<!-- ## Conclusion -->
<!-- - gnagnagni -->
